Modelling state-switching functional data with hidden Markov models: supplement A (case study)

### 2021-07-09 Report/Sections/intro.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

Biologging technology now provides researchers with kinematic data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers have brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. The study of animal movement in particular has been transformed by tracking devices that record kinematic information in a variety of environments \citep{Borger:2020,Dot:2016b}. Tags can record over 50 observations per second, resulting in time series containing millions of observations over the course of several hours. %These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians and biologists. 
These data contain a wealth of information about human and animal behaviour, but modelling these large data sets poses a challenge for statisticians and biologists. One particular difficulty is that simultaneous coarse- and fine-scale processes are often reflected in high-frequency data, and each scale can exhibit unique and complicated dependence structures.

@@

Some FDA models account for between-curve dependence that occurs when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time. For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}. %However, these multilevel models are not appropriate in many biologging applications since they do not account for a discrete number curve types. 
However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time. This is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
In addition to multi-level models, FDA researchers have used functional time series to model dependence in a sequence of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by a discrete number of well-defined hidden states.

>>>>

Biologging technology now provides researchers with kinematic data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers have brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. The study of animal movement in particular has been transformed by tracking devices that record kinematic information in a variety of environments \citep{Borger:2020,Dot:2016b}. Tags can record over 50 observations per second, resulting in time series containing millions of observations over the course of several hours. %These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians and biologists. 
These data contain a wealth of information about human and animal behaviour, but modelling these large data sets poses a challenge for statisticians and biologists. One particular difficulty is that simultaneous coarse- and fine-scale processes are often reflected in high-frequency data, and each scale can exhibit unique and complicated structure.

@@

Some FDA models account for between-curve dependence that occurs when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time. For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}. %However, these multilevel models are not appropriate in many biologging applications since they do not account for a discrete number curve types. 
However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time. This is not appropriate in many biologging applications where behaviours often change suddenly between a discrete number of types.
In addition to multi-level models, FDA researchers have used functional time series to model dependence in a sequence of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by a discrete number of well-defined hidden states.

### 2021-07-09 Report/Sections/models.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

Consider a sequence of $T$ curves, where curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$. Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of fine-scale quantities derived from curve $t$ and indexed by $t^*$. Both $Y_t$ and $Y^*_{t,t^*}$ can be either vectors or scalars. We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$. We assume that the curves $Y_1,\ldots,Y_T$ are indexed according to the temporal order in which they are observed, but the $T$ observations need not be equally spaced in time. To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure. We then show how each of these generalized HMMs can be synthesized to form a wide variety of more complicated models.

@@

\begin{equation}
    \calL_{\text{HMM}}(x,\theta,\Gamma;y) = \delta_{x_1} f^{(x_1)}(y_1; \theta^{(x_1)}) \prod_{t=2}^T \Gamma_{x_{t-1} x_t} f^{(x_t)}(y_t; \theta^{(x_t)}),
    \label{eqn:like}
\end{equation}
%
but it is also computationally tractable to sum over all possible values of $x$ to obtain the following:

@@

It is important to note that replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation. Readers interested in summary statistics that prediction error for sequential functional data can read \citet{Aue:2015} and \citet{Gao:2019}, who both use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction. While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.

>>>>

Consider a sequence of $T$ curves, where curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$. Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of fine-scale quantities derived from curve $t$ and indexed by $t^*$. Both $Y_t$ and $Y^*_{t,t^*}$ can be either vectors or scalars. We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$. We assume that the curves $Y_1,\ldots,Y_T$ are indexed according to the order in which they are observed in time, but the $T$ observations need not be equally spaced apart. To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure. We then show how each of these generalized HMMs can be synthesized to form a wide variety of more complicated models.

@@

\begin{equation}
    \calL_{\text{HMM}}(x,\theta,\Gamma;y) = \delta_{x_1} f^{(x_1)}(y_1; \theta^{(x_1)}) \prod_{t=2}^T \Gamma_{x_{t-1} x_t} f^{(x_t)}(y_t; \theta^{(x_t)}),
    \label{eqn:like}
\end{equation}
%
but it is also tractable to sum over all possible values of $x$ to obtain the following:

@@

Replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation. While we are primarily interested in classifying and describing coarse- and fine-scale behaviour, some researchers may be less interested in interpretability and more interested in predicting future functional data. To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction. While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.

### 2021-07-09 Report/Sections/results.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

%To illustrate an application of this method and compare the candidate models, 
We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{SciPy} package in Python \citep{Virtanen:2019}. Each model is fit 100 times using random initializations, and
%Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation. If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values. If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values. 
we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}. Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory. Most, but not all, initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses. Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of supplement A.

>>>>

%To illustrate an application of this method and compare the candidate models, 
We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the SciPy package in Python \citep{Virtanen:2019}. Each model is fit 100 times using random initializations, and
%Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation. If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values. If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values. 
we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}. Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory. Most, but not all, initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses. Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of supplement A.

### 2021-07-09 Report/Supp_Case.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$. All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
        \addtocounter{fignum}{1}

        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale probability transition matrix parameters $\eta_{1,2}$ and $\eta_{2,1}$. Recall that $\Gamma_{i,j} = \exp(\eta_{i,j}) / \sum_k \exp(\eta_{i,k})$. All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \eta_{1,2},\hat \eta_{2,1})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the wiggliness parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        % eta1
        \begin{center}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_0.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_0.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_1.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_1.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_2.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\eta^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$). Recall that $\eta^{*(i)}_{i^*,i^*} = 0$ and that $\Gamma^{*(i)}_{i^*,j^*} = \exp(\eta^{*(i)}_{i^*,j^*}) / \sum_{k^*} \exp(\eta^{*(i)}_{i^*,k^*})$. All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by a red star. Note that the long ridges of approximately equal likelihood are due to the fact that $\hat \Gamma^{*(i)}_{1,3} \approx \hat \Gamma^{*(i)}_{3,1} \approx 0$. This indicates that the whale rarely switches directly between subdive states 1 and 3.
        \addtocounter{fignum}{1}

>>>>

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$. All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
        \addtocounter{fignum}{1}

        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-gamma-likelihood.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale probability transition matrix parameters $\Gamma_{1,2}$ and $\Gamma_{2,1}$. All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \Gamma_{1,2},\hat \Gamma_{2,1})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the wiggliness parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$. All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
        \addtocounter{fignum}{1}

@@

        % eta1
        \begin{center}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-0-row_0.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-1-row_0.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-0-row_1.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-1-row_1.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-0-row_2.png}
        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-1-row_2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\Gamma^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$). All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1. The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by a red star. Note that $\hat \Gamma^{*(i)}_{1,3} \approx \hat \Gamma^{*(i)}_{3,1} \approx 0$. This indicates that the whale rarely switches directly between subdive states 1 and 3.
        \addtocounter{fignum}{1}
